Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-02-14 paper/content/introduction.tex
Intro: some edits in "results".

Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for making decisions under uncertainty and trading off exploration and exploitation (making good near-term decisions). While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithms with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth.

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that performs better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.

Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of the ``inverted-U relationship" between competition and innovation (see Figure~\ref{fig:inverted-U}). This relationship -- too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better -- is a familiar theme in the economics literature, dating back to \cite{Schumpeter-42}. However, the ``inverted-U relationship'' is driven by different aspects in our model than the ones in the existing literature in economics. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, as opposed to the R\&D costs (which is the more standard cause in prior work).

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate these two effects. We find that the data advantage alone can lead to a significant advantage under competition. In fact, even a small ``head start'' can get amplified into a large incumbency advantage under competition. This effect is most pronounced when the incumbent commits to a more advanced bandit algorithm. These findings suggest that competition dynamics are pertinent to the anti-trust considerations of data as a barrier to entry.

>>>>

Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for making decisions under uncertainty and trading off exploration and exploitation (making good near-term decisions). While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithm with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. \asedit{Taken to the extreme, such dynamics may create a ``death spiral" effect when the vast majority of customers eventually switch to the competitors.}

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm. As a result, consumer welfare is highest under temporary monopoly. \asedit{We find strong evidence of the ``death spiral" effect mentioned earlier; this effect is strongest under permanent duopoly.}

%and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.\swcomment{why ``even''?}

Interpreting the adoption of better algorithms as ``innovation", our
findings can be framed in terms of the ``inverted-U relationship"
between competition and innovation (see Figure~\ref{fig:inverted-U}).
This is a well-established concept in the economics literature, dating
  back to \cite{Schumpeter-42}, whereby too
  little or too much competition is bad for innovation, but
  intermediate levels of competition tend to be better. However, the
``inverted-U relationship'' is driven by different aspects in our
model than the ones in the existing literature in economics. In our
case, the barriers for innovations arise entirely from the
reputational consequences of exploration in competition, as opposed to
the R\&D costs (which is the more standard cause in prior work).

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}
      \draw[->] (-.5,0) -- (7,0) node[right] {};
      \draw[->] (0,-.5) -- (0,3) node[above] {Better algorithms};
      \draw[scale=0.6,domain=0.7:9.8,smooth,variable=\x,blue, line width=0.3mm] plot ({\x},{4.5 - 0.18 * (\x - 5.25)^2});
     \node[below] at (0.7, -0.22) {\footnotesize monopoly};
     \node[below] at (3, -0.2) {\footnotesize incumbent};
     \node[below] at (4.5, -0.22) {\footnotesize duopoly};
     \node[below] at (6, -0.2) {\footnotesize entrant};
 \end{tikzpicture}
 \caption{A stylized ``inverted-U relationship" between strength of competition and ``level of innovation".}
\label{fig:inverted-U}
\end{center}
\end{figure}

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate and compare these two effects. We find that either effects alone leads to a significant advantage under competition. The data advantage is larger than reputation advantage when the incumbent commits to a more advanced bandit algorithm.

\asedit{Data advantage is significant from the anti-trust perspective, as a possible barrier to entry. We find that even a small amount ``data advantage" gets amplified under competition, causing a large difference in eventual market shares. This observation runs contrary to prior work  \cite{varian2018artificial,lambrecht2015can,bajari2018impact}, which studied learning without competition, and found that small amounts of additional data do not provide significant improvement in eventual outcomes. We conclude that competition dynamics -- that firms compete as they learn over time -- are pertinent to these anti-trust considerations.}
